---
title: "Adding a new TI method"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Adding a new TI method}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---

This vignette describes how to include your own method into the dynverse workflow. We'll illustrate this with a very basic method: using one of the components of a PCA as the pseudotime.

The main function to create a new ti method is `create_ti_method`. This wraps:

* The name of the method
* The main function to run a method on a dataset
* The parameters required by a method, and their ranges

## Input

To test our method, we will use a very simple toy data example:

```{r, message=FALSE}
library(dynwrap)
library(dynmethods)
library(tidyverse)
library(ParamHelpers)
```

The main function will receive three different types of input parameters: the expression data, extra parameters and prior information.

### Expression data

The The raw `counts` or normalised `expression`. This is a matrix with in the columns different features (genes) and in the rows different cells.

```{r}
ncells <- 1000
pseudotime <- runif(ncells)

expression <- matrix(
  c(
    (pseudotime - 0.5) ** 2,
    sqrt(pseudotime + 20),
    pseudotime
  ),
  ncol = 3,
  dimnames = list(as.character(rep(seq_len(ncells))), as.character(c("A", "B", "C")))
)
expression <- expression + rnorm(length(expression), sd = 0.02)
```

In this example, we'll use the expression data as follows:
```{r}
pca <- prcomp(expression)
```

### Parameters

The main function can accept any additional parameters, which should also be defined in a parameter set using the `ParamHelpers` package. This allows you to specify the type of parameter (integer, numeric, discrete, logical, ...), the possible range of values and the default value.

In this example we'll define one parameter, the `component`:

```{r}
par_set <- makeParamSet(
  makeIntegerParam("component", lower = 1, upper = 10, default = 1)
)
```

```{r}
component <- 1
pseudotime <- pca$x[, component]
pseudotime <- (pseudotime - min(pseudotime)) / (max(pseudotime) - min(pseudotime))
```

### Prior information

Prior information can be required (no default) or optional (default is `NULL`). 

Following prior information requests are possible:

```{r, echo=FALSE}
data("priors", package = "dynwrap")
priors %>% 
  select(prior_id, prior_description) %>% 
  rename(`Name of parameter` = prior_id, `Description` = prior_description) %>% 
  knitr::kable("markdown")
```

In this example we will use the `start_cell_id`:
```{r}
start_cell_ids <- as.character(which.min(pseudotime))
```

```{r}
if (!is.null(start_cell_ids)) {
  if(mean(pseudotime[start_cell_ids]) > 0.5) {
    pseudotime <- 1-pseudotime
  }
}
```
## Output

In the end, the output of the method should be transformed into the common trajectory format:

**[Figure of common output model]**

The topology of the trajectory requires:

* `milestone_ids` The ids of the milestones in the trajectory. Type: Character vector.
* `milestone_network` The network of the milestones. Is a dataframe with columns from, to, length and whether it is directed
* `divergence_regions` A data frame specifying the divergence regions between milestones (e.g. a bifurcation). Is a dataframe containing a divergence_id, milestone_id, and is_start (TRUE when this milestone is the start of this divergence

```{r}
milestone_ids <- c("A", "B")
milestone_network <- tibble(from = "A", to = "B", length = 1, directed = TRUE)
divergence_regions <- tibble()
```

The location of the cells can be provided in two ways:

* `milestone_percentages` A data frame specifying what percentage milestone each cell consists of. Is a dataframe with columns cell_id, milestone_id, and percentage
* `progressions` Specifies the progression of a cell along a transition in the milestone_network. Is a dataframe with columns cell_id, from, to, and percentage

Example of `milestone_percentages`:
```{r}
milestone_percentages <- bind_rows(
  tibble(
    milestone_id = "A",
    cell_id = names(pseudotime),
    percentage = 1-pseudotime
  ),
  tibble(
    milestone_id = "B",
    cell_id = names(pseudotime),
    percentage = pseudotime
  )
)
```

Example of `progressions`:
```{r}
progressions <- tibble(
  cell_id = names(pseudotime),
  from = "A",
  to = "B",
  percentage = pseudotime
)
```

We can now wrap this information up:
```{r}
trajectory <- 
  wrap_data(
    cell_id = names(pseudotime)
  ) %>% 
  add_trajectory(
    milestone_ids = milestone_ids,
    milestone_network = milestone_network,
    divergence_regions = divergence_regions,
    progressions = progressions # either milestone_percentages or progressions have to be provided
  )
```

In some cases, we already provide functions which do the calculation of milestone_percentages or progressions. For example, when a method returns only a pseudotime:

```{r}
trajectory <- 
  wrap_data(
    cell_id = names(pseudotime)
  ) %>% 
  add_linear_trajectory(pseudotime)
```

Other examples include:

**[Figure of common wrapping methods]**

We can plot this output using the plotting functions provided by the `dynplot` package:

```{r}
dynplot::plot_dimred(trajectory, "pseudotime", expression_source = expression)
```


## Creating the method object

Combining everything in a function:

```{r}
run_fun <- function(expression, component, start_cell_id = NULL) {
  pca <- prcomp(expression)
  
  pseudotimes <- pca$x[, component]
  
  milestone_ids <- c("A", "B")
  milestone_network <- tibble(from="A", to="B", length=1, directed=TRUE)
  divergence_regions <- tibble()
  
  progressions <- tibble(
    cell_id = names(pseudotime),
    from = "A",
    to = "B",
    percentage = pseudotime
  )
  
  trajectory <- 
    wrap_data(
      cell_id = names(pseudotime)
    ) %>% 
    add_trajectory(
      milestone_ids = milestone_ids,
      milestone_network = milestone_network,
      divergence_regions = divergence_regions,
      progressions = progressions # either milestone_percentages or progressions have to be provided
    )
}
```

We can now create the TI object

```{r}
ti_dummy <- create_ti_method(
  "dummy", 
  par_set,
  run_fun
)
```

This function also has several alternative parameters, among which:

* R packages which have to be installed, and R packages which have to be installed and loaded
* The function to plot the output of a method, default is `dynplot::plot_default`
* A short name of the method

## Running the method

Running the method is easy given a task.

```{r}
task <- wrap_data("", rownames(expression)) %>% add_expression(expression, expression)

model <- infer_trajectory(task, ti_dummy())
```


```{r}
library(dynplot)
dynplot::plot_dimred(model, color_cells = "pseudotime" , expression_source = task$expression)
```

